Improving the Accuracy of a Sequential Mining Model for Hurricane Trajectory Prediction

### 2018-01-29 paper/manuscript_revised_jan25.tex
Fix repo structure. Clarify uniqueness of weighted data contribution

\maketitle

% Reduce repition: highlight modificatiosn of aprioriall less frequently
% Clearly state that weighted support is our idea, inspired by some work but we have a diff approach

\begin{abstract}

In the 21st century alone, hurricanes have caused billions of dollars worth of damage due to property destruction, personal danger, and civil unrest. The catastrophic impact of hurricanes and other tropical storms has motivated extensive research into hurricane forecasting. Hurricane trajectory prediction is a subfield of hurricane forecasting that traditionally uses meteorological analysis to determine the future path of a given hurricane. While meteorological methods of hurricane trajectory prediction have been effective in the past, little attention has been paid to hurricane trajectory prediction with respect to historical frequent pattern mining. In this work, we propose multiple dimensions for improving the accuracy of data mining-based hurricane trajectory prediction. We propose the use of the Haversine distance formula for realistic hurricane trajectory distance comparison. We also propose the use of weighted training data to favour recent hurricane trends when determining a trajectory prediction. Finally, we use training data exclusively from the years 1950 to 2000 to reduce the impact of historical climate trends and inaccuracies in hurricane recording methodologies. After implementing all three improvement dimensions on a contemporary data mining model for hurricane trajectory prediction based on AprioriAll, results show a prediction-correctness ratio up to 10.0\% higher than the current state-of-the-art, at 75.0\%.

@@

To the knowledge of the authors, the existing standard in hurricane trajectory prediction via data mining is established by Dong et al in \cite{major-hurricane-model}. The work cited proposes a hurricane trajectory prediction model based on data mining (HTPDM) that uses a modified variant of AprioriAll, a common sequential mining algorithm \cite{AprioriAll-original}. HTPDM executes AprioriAll against historical hurricane trajectories from 1900 to 2000 to generate interesting trajectory association rules. These association rules are compared against test data composed of hurricane trajectories from 2001 to 2008. HTPDM divides a given testing trajectory into two parts: the initial trajectory $T_{i}$ and the terminal trajectory $T_{t}$. $T_{i}$ is compared against all association rule antecedents until a sufficiently matching rule R is found based on a predetermined fitness function. The consequent of R is then compared against $T_{t}$ using pattern-matching techniques to determine if the prediction was correct. The results of HTPM showed that the model was able to achieve a best-case correctness ratio of 65.0\%. In this work, a model similar to HTPDM will be implemented and various strategies will be employed to increase the correctness ratio. Reducing the impact of historical factors, favouring the impact of recent trends, and improving the realism of the hurricane trajectory prediction model result in a new model dubbed WARD-HTP (Weighted-Asset, Realistic-Distance Hurricane Trajectory Predictor). The strategies employed in WARD-HTP ultimately result in a correctness ratio of 75.0\%, a 10\% improvement over HTPDM.

@@

The second approach to increasing trajectory prediction correctness deployed in WARD-HTP is the usage of weighted training data in AprioriAll. Applying weights to recorded trajectories in chronological order allows AprioriAll to favour recently produced trajectory patterns by increasing the support and confidence of such recently produced trajectories. While the weighting of training data results in some trajectories being counted with a support greater than 1, this effect is balanced by counting earlier recorded trajectories with a support less than 1. In WARD-HTP, we weigh hurricane trajectories on a scale increasing linearly. This results in a tipping point \textit{T} where records before \textit{T} are underweighted and records after \textit{T} are overweighted. Depending on the weighing approach used, the combined weight of all records does not necessarily need to be balanced.

The following algorithm applies further modifications to AprioriAll in order to weigh the stored hurricane records. In this algorithm, W identifies the total weighing scale to be used in the modification. For example, if W$=$3, then the final recorded hurricane trajectory in the hurricane database HDB will be counted with a support of 3. In addition, the first trajectory will be counted with a support of 3/$\lvert$HDB$\rvert$. The support modifier increases by 3/$\lvert$HDB$\rvert$ for each entry in HDB.

@@

The algorithm described assumes that the HDB is ordered chronologically. This ensures that recently produced hurricane trajectory records correspond to a higher weight. Iteration through the HDB increases the current weight w by the weighting scale W divided by the size of the HDB. Therefore the final record in the HDB is guaranteed to have a weighted support exactly equal to W. Alternatively, the support weight modifier could also change every year. This means that W would be divided by the number of years which the database records span, and the value of w would increase every time the records reach a new year. Though index-based weighting is shown above in algorithm (3), both weighting strategies will be evaluated further in this work.

Under the assumption of chronologically ordered data, introducing weighted support to AprioriAll will result in higher support for patterns which occur in recent years. Fewer repeated records will be needed to generate frequent sequences from recent data, and more repeated records will be needed to generate frequent sequence from older data. Therefore, the introduction of weighted support to AprioriAll is expected to further increase the hurricane trajectory prediction correctness ratio for current and future trajectories being tested upon.

@@

\begin{thebibliography}{1}
\bibitem{hurricane-cost} Pielke Jr, Roger A., et al. Normalized hurricane damage in the United States: 1900–2005. Natural Hazards Review 9.1 (2008): 29-42
\bibitem{hurricane-cost-non-financial} F. F. Townsend, The Federal Response to Hurricane Katrina: Lessons Learned, United States Federal Government, Washington, DC, USA, 2006
\bibitem{data mining-def} S. Sumathi and S. N. Sivanandam, Introduction to Dataming and its Applications: Springer Publishing 2006, ISBN: 978-3-540-34350-9
\bibitem{major-hurricane-model} Dong, X and C. Pi, D. (2013). Novel method for hurricane trajectory prediction based on data mining. Natural Hazards and Earth System Sciences. 13. 3211-3220. 10.5194/nhess-13-3211-2013. 
\bibitem{HURDAT2-original} C. Landsea, J. Franklin and J. Beven (2015) The revised Atlantic hurricane database (HURDAT2). National Oceanic and Atmospheric Administration
\bibitem{AprioriAll-original} R. Agrawal and R. Srikant, Mining sequential patterns, Proceedings of the Eleventh International Conference on Data Engineering, Taipei, 1995, pp. 3-14.
doi: 10.1109/ICDE.1995.380415
\bibitem{hurricane-intensity-1} Y. Su, S. Chelluboina, M. Hahsler and M. H. Dunham, A New Data Mining Model for Hurricane Intensity Prediction, 2010 IEEE International Conference on Data Mining Workshops, Sydney, NSW, 2010, pp. 98-105
doi: 10.1109/ICDMW.2010.158
\bibitem{hurricane-intensity-2} R. Yang, J. Tang and D. Sun (2011), Association Rule Mining Application for Atlantic Tropical Cyclone Intensity Changes, American Meteorological Society. doi: 10.1175/WAF-D-10-05029.1
\bibitem{hurricane-forecasting-1} Rozanova, Olga S., Jui-Ling Yu, and Chin-Kun Hu. Typhoon eye trajectory based on a mathematical model: Comparing with observational data. Nonlinear Analysis: Real World Applications 11.3 (2010): 1847-1861.
\bibitem{hurricane-forecasting-2} Chandan Roy and Rita Kovordanyi, Tropical cyclone track forecasting techniques: A review, 2012, Atmospheric research, (104-105), 40-69. doi: 10.1016/j.atmosres.2011.09.012.
\bibitem{typhoon-clustering} Kim, Hyeong-Seog, et al. Pattern classification of typhoon tracks using the fuzzy c-means clustering method. Journal of Climate 24.2 (2011): 488-508.
\bibitem{matplotlib} J. D. Hunter, Matplotlib: A 2D Graphics Environment, in Computing in Science and Engineering, vol. 9, no. 3, pp. 90-95, May-June 2007.
doi: 10.1109/MCSE.2007.55
\bibitem{lat-long-distance} Menno-Jan, K. (2013). Map use: reading, analysis, interpretation: by Kimberling, AJ, Buckley, AR, Muehrcke, PC and Muehrcke, JO, Redlands, California, ESRI Press Academic, 2012. ISBN 978-1-589-48279-1 Seventh edition.
\bibitem{hurricane-distances} D. Brown. (2013) Forecast Verification: Quantifying Forecast Uncertainty. National Hurricane Conference: Hurricane Readiness for Coastal Communities
\bibitem{earth-radius-kilometres} Chambat, F., and Valette, B. (2001). Mean radius, mass, and inertia for reference Earth models. Physics of the Earth and Planetary Interiors, 124(3), 237-253.
\bibitem{hurricane-recording-history} Landsea, C. W., Vecchi, G. A., Bengtsson, L., and Knutson, T. R. (2010). Impact of duration thresholds on Atlantic tropical cyclone counts. Journal of Climate, 23(10), 2508-2519.

>>>>

\maketitle
% Clearly state that weighted support is our idea, inspired by some work but we have a diff approach
\begin{abstract}

In the 21st century alone, hurricanes have caused billions of dollars worth of damage due to property destruction, personal danger, and civil unrest. The catastrophic impact of hurricanes and other tropical storms has motivated extensive research into hurricane forecasting. Hurricane trajectory prediction is a subfield of hurricane forecasting that traditionally uses meteorological analysis to determine the future path of a given hurricane. While meteorological methods of hurricane trajectory prediction have been effective in the past, little attention has been paid to hurricane trajectory prediction with respect to historical frequent pattern mining. In this work, we present multiple dimensions for improving the accuracy of data mining-based hurricane trajectory prediction. We first propose the use of the Haversine distance formula for realistic hurricane trajectory distance comparison. We also introduce the use of weighted training data to favour recent hurricane trends when determining a trajectory prediction. Finally, we use training data exclusively from the years 1950 to 2000 to reduce the impact of historical climate trends and inaccuracies in hurricane recording methodologies. After implementing all three improvement dimensions on a contemporary data mining model for hurricane trajectory prediction based on AprioriAll, results show a prediction-correctness ratio up to 10.0\% higher than the current state-of-the-art, at 75.0\%.

@@

The existing standard in hurricane trajectory prediction via association rule mining is established by Dong et al in \cite{major-hurricane-model}. The work cited proposes a hurricane trajectory prediction model based on data mining (HTPDM) that uses a modified variant of AprioriAll, a standard algorithm for mining sequential data \cite{AprioriAll-original}. HTPDM executes AprioriAll against historical hurricane trajectories from 1900 to 2000 to generate interesting trajectory association rules. These association rules are compared against test data composed of hurricane trajectories from 2001 to 2008. HTPDM divides a given testing trajectory into two parts: the initial trajectory $T_{i}$ and the terminal trajectory $T_{t}$. $T_{i}$ is compared against all association rule antecedents until a sufficiently matching rule R is found based on a predetermined fitness function. The consequent of R is then compared against $T_{t}$ using pattern-matching techniques to determine if the prediction was correct. The results of HTPM showed that the model was able to achieve a best-case correctness ratio of 65.0\%. In this work, a model similar to HTPDM will be implemented and various strategies will be employed to increase the correctness ratio. Reducing the impact of historical factors, favouring the impact of recent trends, and improving the realism of the hurricane trajectory prediction model result in a new model dubbed WARD-HTP (Weighted-Asset, Realistic-Distance Hurricane Trajectory Predictor). The strategies employed in WARD-HTP ultimately result in a correctness ratio of 75.0\%, a 10\% improvement over HTPDM.

@@

The second approach to increasing trajectory prediction correctness deployed in WARD-HTP is the use of weighted training data in AprioriAll. Applying chronologically-ordered weights to recorded trajectories allows AprioriAll to favour recently produced trajectory patterns by increasing their support. While the weighting of training data results in some trajectories being counted with a support greater than 1, this effect is balanced by counting earlier recorded trajectories with a support less than 1. However, the most promising results of this work were produced when the support values for recent trajectories were increased \textit{more} than the support values for early trajectories were decreased.

In WARD-HTP, we weigh hurricane trajectories on a linearly increasing scale. This results in a tipping point \textit{T} where records before \textit{T} are underweighted and records after \textit{T} are overweighted. Depending on the weighing approach used, the combined weight of all records does not necessarily need to equal 1. The following algorithm applies further modifications to AprioriAll in order to weigh the stored hurricane records. In this algorithm, W identifies the total weighing scale to be used in the modification. For example, if W$=$3, then the final recorded hurricane trajectory in the hurricane database HDB will be counted with a support of 3. In addition, the first trajectory will be counted with a support of 3/$\lvert$HDB$\rvert$. The support modifier increases by 3/$\lvert$HDB$\rvert$ for each entry in HDB.

@@

The algorithm described assumes that the HDB is ordered chronologically. This ensures that recently produced hurricane trajectory records correspond to a higher weight. Iteration through the HDB increases the current weight w by the weighting scale W divided by the size of the HDB. Therefore the final record in the HDB is guaranteed to have a weighted support exactly equal to W. Alternatively, the support weight modifier may also be changed year by year. This means that W would be divided by the number of years which the database records span, and the value of w would increase every time the records reach a new year. Though index-based weighting is shown above in algorithm (3), both year and index-based weighting strategies will be evaluated further in this work.

Assuming the input data is ordered chronogically, introducing weighted support to AprioriAll will result in higher support for patterns which occur in recent years. Recent trajectories will require less repitition to produce associaton rules, and higher repitition will be needed to generate frequent sequences from older data. The approach of chronogically weighted data will produce association rules that favor recent trends in hurricane trajectories. Therefore, the introduction of weighted support to AprioriAll is expected to further increase the hurricane trajectory prediction correctness ratio for current and future trajectories being tested upon. Publications such as \cite{war-mining} and \cite{cai-weighted-thesis} first introduce the use of weighted support for association rule mining. However, to the knowledge of the authors this work is the first to (1) apply weighted support to sequential data, (2) weigh individual records based on a proportional scale and (3) allow records to have a weight less than 1.

@@

\begin{thebibliography}{1}
\bibitem{hurricane-cost} Pielke Jr, Roger A., et al. Normalized hurricane damage in the United States: 1900–2005. Natural Hazards Review 9.1 (2008): 29-42
\bibitem{hurricane-cost-non-financial} F. F. Townsend, The Federal Response to Hurricane Katrina: Lessons Learned, United States Federal Government, Washington, DC, USA, 2006
\bibitem{data mining-def} S. Sumathi and S. N. Sivanandam, Introduction to Dataming and its Applications: Springer Publishing 2006, ISBN: 978-3-540-34350-9
\bibitem{major-hurricane-model} Dong, X and C. Pi, D. (2013). Novel method for hurricane trajectory prediction based on data mining. Natural Hazards and Earth System Sciences. 13. 3211-3220. 10.5194/nhess-13-3211-2013. 
\bibitem{HURDAT2-original} C. Landsea, J. Franklin and J. Beven (2015) The revised Atlantic hurricane database (HURDAT2). National Oceanic and Atmospheric Administration
\bibitem{AprioriAll-original} R. Agrawal and R. Srikant, Mining sequential patterns, Proceedings of the Eleventh International Conference on Data Engineering, Taipei, 1995, pp. 3-14.
doi: 10.1109/ICDE.1995.380415
\bibitem{hurricane-intensity-1} Y. Su, S. Chelluboina, M. Hahsler and M. H. Dunham, A New Data Mining Model for Hurricane Intensity Prediction, 2010 IEEE International Conference on Data Mining Workshops, Sydney, NSW, 2010, pp. 98-105
doi: 10.1109/ICDMW.2010.158
\bibitem{hurricane-intensity-2} R. Yang, J. Tang and D. Sun (2011), Association Rule Mining Application for Atlantic Tropical Cyclone Intensity Changes, American Meteorological Society. doi: 10.1175/WAF-D-10-05029.1
\bibitem{hurricane-forecasting-1} Rozanova, Olga S., Jui-Ling Yu, and Chin-Kun Hu. Typhoon eye trajectory based on a mathematical model: Comparing with observational data. Nonlinear Analysis: Real World Applications 11.3 (2010): 1847-1861.
\bibitem{hurricane-forecasting-2} Chandan Roy and Rita Kovordanyi, Tropical cyclone track forecasting techniques: A review, 2012, Atmospheric research, (104-105), 40-69. doi: 10.1016/j.atmosres.2011.09.012.
\bibitem{typhoon-clustering} Kim, Hyeong-Seog, et al. Pattern classification of typhoon tracks using the fuzzy c-means clustering method. Journal of Climate 24.2 (2011): 488-508.
\bibitem{matplotlib} J. D. Hunter, Matplotlib: A 2D Graphics Environment, in Computing in Science and Engineering, vol. 9, no. 3, pp. 90-95, May-June 2007.
doi: 10.1109/MCSE.2007.55
\bibitem{lat-long-distance} Menno-Jan, K. (2013). Map use: reading, analysis, interpretation: by Kimberling, AJ, Buckley, AR, Muehrcke, PC and Muehrcke, JO, Redlands, California, ESRI Press Academic, 2012. ISBN 978-1-589-48279-1 Seventh edition.
\bibitem{hurricane-distances} D. Brown. (2013) Forecast Verification: Quantifying Forecast Uncertainty. National Hurricane Conference: Hurricane Readiness for Coastal Communities
\bibitem{earth-radius-kilometres} Chambat, F., and Valette, B. (2001). Mean radius, mass, and inertia for reference Earth models. Physics of the Earth and Planetary Interiors, 124(3), 237-253.
\bibitem{war-mining} W. Wang, J. Yang and P. Yu Efficient  mining of  weighted association rules (WAR), Proc. of the ACM SIGKDD Conf. on Knowledge Discovery and Data Mining, 270-274, 2000.
\bibitem{cai-weighted-thesis} Cai, Chun Hing, Ada Wai-Chee Fu, C. H. Cheng, and W. W. Kwong. Mining association rules with weighted items. In Database Engineering and Applications Symposium, 1998. Proceedings. IDEAS'98. International, pp. 68-77. IEEE, 1998.
\bibitem{hurricane-recording-history} Landsea, C. W., Vecchi, G. A., Bengtsson, L., and Knutson, T. R. (2010). Impact of duration thresholds on Atlantic tropical cyclone counts. Journal of Climate, 23(10), 2508-2519.
